---
title: "Homework 1 - readable and efficient R code"
author: "Shuang Du"
date: "`r format(Sys.time(), '%m/%d/%Y')`"
output: rmarkdown::github_document
---

# Question 1 - "function-alize" this code

Read over the code below and perform the following:

* Wrap it into a function `foobar0` which has arguments `x` and `z`
  and which returns the vector `x` at the end of the following code.
* Rewrite this into a function `foobar` which is easier to read, by
  reducing repetitive code. E.g. `foobar` might call a function to
  check the input, and another function to perform the three lines of
  computation. 
* Check that the two versions produce the same output using the
  function `all.equal`.

```{r}
set.seed(1)
x <- rnorm(100)
z <- rnorm(100)
if (sum(x >= .001) < 1) {
  stop("step 1 requires 1 observation(s) with value >= .001")
}
fit <- lm(x ~ z)
r <- fit$residuals
x <- sin(r) + .01
if (sum(x >= .002) < 2) {
  stop("step 2 requires 2 observation(s) with value >= .002")
}
fit <- lm(x ~ z)
r <- fit$residuals
x <- 2 * sin(r) + .02
if (sum(x >= .003) < 3) {
  stop("step 3 requires 3 observation(s) with value >= .003")
}
fit <- lm(x ~ z)
r <- fit$residuals
x <- 3 * sin(r) + .03
if (sum(x >= .004) < 4) {
  stop("step 4 requires 4 observation(s) with value >= .004")
}
fit <- lm(x ~ z)
r <- fit$residuals
x <- 4 * sin(r) + .04
x
```
The function that can wrap up the above calculation is:

```{r}
footbar0<- function(x,z){
  if (sum(x >= .001) < 1) {
    stop("step 1 requires 1 observation(s) with value >= .001")
  }
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- sin(r) + .01
  if (sum(x >= .002) < 2) {
    stop("step 2 requires 2 observation(s) with value >= .002")
  }
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- 2 * sin(r) + .02
  if (sum(x >= .003) < 3) {
    stop("step 3 requires 3 observation(s) with value >= .003")
  }
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- 3 * sin(r) + .03
  if (sum(x >= .004) < 4) {
    stop("step 4 requires 4 observation(s) with value >= .004")
  }
  fit <- lm(x ~ z)
  r <- fit$residuals
  x <- 4 * sin(r) + .04
  return (x)
}

```

A more efficient function can be written as:
```{r}
# Help check the condition to proceed
fooCheck <- function(x, i){
  if (sum(x >= (.001*i)) < i) {
  stop(paste0("step ", i, " requires ", i, " observation(s) with value >= ", i*.001))
}
}

# Perform the three lines of computation
fooComp <- function(x, z, i){
  x <- i * sin(lm(x ~ z)$residuals) + .01 * i
  return(x)
}

# A new version of the function which gets rid of repetitive sections
footbar <- function(x, z){
  for (i in 1:4){
    fooCheck(x, i)
    x <- fooComp(x, z, i)
  }
  return(x)
}
```

Other attemps:
```{r}
#attemp1
footbar1 <- function(x, z) {
  lapply(seq_len(4), function(i) (i * sin(lm(x ~ z)$residuals) + .01 * i))
  return (x)
}
```


```{r}
#attemp2
footbar2 <- function(x, z) {
  for (i in 1:4) {
    x <- i * sin(lm(x ~ z)$residuals) + .01 * i
    if (sum(x >= .001 * (i)) < i) {
      stop("step",i, "requires",i, "observation(s) with value >= .001")
    }
  }
  return (x)
}
```


```{r}
library(microbenchmark)
all.equal(footbar0(x,z),footbar(x,z))
microbenchmark(footbar0(x,z),footbar(x,z),footbar1(x,z),footbar2(x,z))
```
### Conclusion: The new function `footbar(x,z)` has get rid of repetitive codes and has the same output as the original function `footbar0(x,z)`.


# Question 2 - vectorize this code and benchmark

* Take the following function `f0` and rewrite it as a function `f`,
  which is faster and easier to read, by removing the loop of `i` from
  1 to `m`.
* Benchmark `f` and `f0` using `microbenchmark`. How much faster is `f`?

```{r}
n <- 30
p <- 50
p2 <- 25
m <- 1000
set.seed(1)
x <- matrix(rnorm(n*p),nrow=n,ncol=p)
b <- matrix(rnorm(m*p),nrow=m,ncol=p)
a <- matrix(rnorm(m*p2),nrow=m,ncol=p2)

f0 <- function(x,b,a) {
  out <- numeric(0)
  for (i in seq_len(m)) {
    bb <- b[i,]
    aa <- a[i,]
    out <- c(out, sum(x %*% bb) + sum(aa))
  }
  return (out)
}
```

```{r}
library(microbenchmark)
#attemp1
f00<- function(x,b,a) {
  re <- sapply(seq_len(m), function(i) (sum(x %*% b[i,]) + sum(a[i,])))
  unname(re)
}
```

```{r}
#attemp2
f1 <- function(x,b,a) {
  ans <- numeric(length(m))
  for (i in seq_len(m)) {
    ans[i] <- sum(x %*% b[i,]) + sum(a[i,])
  }
  ans
}

```

```{r}
#attemp3
s = rowSums(a)
f2 <- function(x,b,a){
  y <- function(t){
    x %*% t
  }
  sum(apply(b,1,y)) + s
} 
```

```{r}
# attemp4 
f <- function(x, b, a){
  colSums(x %*% t(b))+rowSums(a)
}
```

```{r}
microbenchmark(f0(x,b,a), f00(x,b,a),f1(x,b,a),f(x,b,a))
```
### Conclusion: function f(x,b,a) has the best performance on runtime, saved more than 50% per cent of original runtime.

# Question 3 - build a faster t-test

* Rewrite the following function `getT0` which computes `m`
  two-sample t-tests (equal variance) between two groups as a function
  `getT`, which is faster by using vectorized operations over the `m`
  sets of observations. (There are functions in R packages, such as
  `genefilter::rowttests` which will quickly perform this operation,
  but I want you to write your own function using simple R functions
  like `rowSums`, etc.)
* Benchmark `getT` and `getT0`. How much faster is `getT`?

```{r}
m <- 400
n <- 50
little.n <- n/2
set.seed(1)
x <- matrix(rnorm(m*n),nrow=m,ncol=n)
f <- gl(2,little.n)

getT0 <- function(x, f) {
  ts <- sapply(seq_len(m), function(i) t.test(x[i,] ~ f, var.equal=TRUE)$statistic)
  unname(ts)
}
```


```{r}
# calculate the variance for each row in a matrix
rowVars <- function(x) {
  rowSums((x-rowMeans(x))^2) / (ncol(x)-1)
}

# A faster t test function
getT <- function (x, f){
  xbar.diff <- rowMeans(x[, which(f==1)]) - rowMeans(x[, which(f==2)]) # calculate the difference in row means
  n1 <- sum(f==1) # sample size in group 1
  n2 <- sum(f==2) # sample size in group 2
  sp.sq <- ((n1-1)*rowVars(x[, which(f==1)])+(n2-1)*rowVars(x[, which(f==2)]))/(n1 + n2 - 2) #s_p squared
  ts <- xbar.diff/sqrt(sp.sq*(1/n1 + 1/n2)) # output a vector of t test statistics
  return(ts)
}
```

```{r}
all.equal(getT0(x,f),getT(x,f))
microbenchmark(getT0(x,f),getT(x,f))
```

### Conclusion: By rewrting the function, `getT(x,f)` is more efficieny than the original function.